
You can do this on your own data, or on included data here.


##Continuous data

```{r, eval=T}
library(geiger)
library(ape)
tree.primates <- read.tree(text="((((Homo:0.21,Pongo:0.21):0.28,Macaca:0.49):0.13,Ateles:0.62):0.38,Galago:1.00);") #using examples from ape ?pic
X <- c(4.09434, 3.61092, 2.37024, 2.02815, -1.46968)
Y <- c(4.74493, 3.33220, 3.36730, 2.89037, 2.30259)
names(X) <- names(Y) <- c("Homo", "Pongo", "Macaca", "Ateles", "Galago")
pic.X <- pic(X, tree.primates)
pic.Y <- pic(Y, tree.primates)
```

Now, positivitize the contrasts and do a regression through the origin.

```{r,eval=T}
pic.X <- abs(pic.X)
pic.y <- abs(pic.Y)
reg <- lm(pic.Y ~ 0 + pic.X)
print(reg)
```


##Discrete data

```{r, eval=T}
library(corHMM)
?corHMM
data(primates)
ls()
print(primates)
require(phytools)
```


Just to make this a better dataset for our exercise, I'm going to change some of the states (I want to have all four trait combinations present).
For actual analyses, of course, DO NOT INVENT YOUR DATA.

First, a review of discrete state models:

```{r, eval=T}
primates$trait[which(grepl("Hylobates",primates$trait[,1])),2]<-1

trait1<-primates$trait[,2]
names(trait1)<-primates$trait[,1]
str(primates)
plotSimmap(make.simmap(primates$tree, trait1), pts=FALSE, fsize=0.8)
rate.mat.er<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
print(rate.mat.er)
```

What does this matrix mean? This matrix represents the indices of the rate parameters to be optimized via corHMM. In this case, it would be the rates from 0->1 and 1->0. As both parameters are indexed at 1, this means the transition rate will be the same, regardless of the starting state.

```{r, eval=T}
pp.er<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(pp.er)
```

What do these results mean? These results mean the rate of going from 0 to 1 is the same as going from 1 to 0.

```{r, eval=T}
rate.mat.ard<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ARD")
print(rate.mat.ard)
```

And these? This indicates the rate parameters can differ.

```{r, eval=T}
pp.ard<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard,node.states="marginal")
print(pp.ard)
```

which model is better? The ER model has a lower AIC and AICc score, so it is the better model.

Now let's look at multiple traits.

This is a matrix with four states

```{r, eval=T}
rate.mat.er.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rate.mat.er.4state)
```

Convert the two binary traits into a single four character state

```{r, eval=T}
fourstate.trait<-rep(NA,Ntip(primates$tree))
for(i in sequence(Ntip(primates$tree))) {
	if(primates$trait[i,2]==0 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-0
	}	
	if(primates$trait[i,2]==0 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-1
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-2
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-3
	}	
}
fourstate.data<-data.frame(Genus_sp=primates$trait[,1], T1=fourstate.trait)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, model="ER", node.states="marginal"))
print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat=rate.mat.er.4state, node.states="marginal", model="ARD"))
rate.mat.ard.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")
print(rate.mat.ard.4state)
```

Now let's make the equivalent of a GTR matrix:

```{r, eval=T}
rate.mat.gtr.4state<-rate.mat.ard.4state
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(1,4))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(2,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(3,8))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(4,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(5,7))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(6,7))
print(rate.mat.gtr.4state)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.gtr.4state, node.states="marginal", model="ARD"))
```


Now make a model like Pagel 1994

```{r, eval=T}
print(rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=2, nstates=2, model="ARD"))
rate.mat.pag94<-rate.par.drop(rate.mat.ard.4state, drop.par=c(3,5,8,10))
print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.pag94, node.states="marginal", model="ARD"))
```

Now that you have some introduction, there are two routes:

##Route 1

**Construct a model to test if state 1 can never be lost**

**Experiment with the effects of frequencies at the root.**

**Create and use a model to see if transitions from 00 go to 11 only via 01.**

```{r,eval=T}
# Drop parameters that lead from state 1 to 0
rate.mat.rt1 <- rate.par.drop(rate.mat.ard.4state,drop.par=c(1,2,3,5,6,8,9))
print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.rt1, node.states="marginal", model="ARD"))

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.rt1, node.states="marginal", model="ARD",root.p = c(0.5,0.25,0.25,0)))

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.rt1, node.states="marginal", model="ARD",root.p = c(0.25,0.5,0.25,0)))

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.rt1, node.states="marginal", model="ARD",root.p = c(0.7,0.1,0.1,0.1)))
# Drop parameters allowing (0,0) and (1,0) to go to (1,1)
rate.mat.01 <- rate.par.drop(rate.mat.ard.4state,drop.par=c(10,12))
print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.01, node.states="marginal", model="ARD"))
```


##Route 2

Maddison and FitzJohn (2015) pretty convincingly show (to me) that Pagel (1994) is just not a good method. Ok. So work on a fix. They point to [Read and Nee (1995)](http://dx.doi.org/10.1006/jtbi.1995.0047) as a low power but possible solution. Look at their appendix, especially, and write an implementation.
