
You can do this on your own data, or on included data here.


##Continuous data

```{r, eval=FALSE}
library(geiger)
library(ape)
#library(pic) #?????????????? 
tree.primates <- read.tree(text="((((Homo:0.21,Pongo:0.21):0.28,Macaca:0.49):0.13,Ateles:0.62):0.38,Galago:1.00);") #using examples from ape 
X <- c(4.09434, 3.61092, 2.37024, 2.02815, -1.46968)
Y <- c(4.74493, 3.33220, 3.36730, 2.89037, 2.30259)
names(X) <- names(Y) <- c("Homo", "Pongo", "Macaca", "Ateles", "Galago")
pic.X <- pic(X, tree.primates)
pic.Y <- pic(Y, tree.primates)
?pic
```

Now, positivitize the contrasts and do a regression through the origin.

##Discrete data

```{r, eval=FALSE}
require("corHMM")
##?corHMM
data(primates)
ls()
print(primates)
plot(primates$tree)
require(phytools)
```


Just to make this a better dataset for our exercise, I'm going to change some of the states (I want to have all four trait combinations present).
For actual analyses, of course, DO NOT INVENT YOUR DATA.

First, a review of discrete state models:

```{r, eval=FALSE}
primates$trait[which(grepl("Hylobates",primates$trait[,1])),2]<-1
trait1<-primates$trait[,2]
names(trait1)<-primates$trait[,1]
plotSimmap(make.simmap(primates$tree, trait1), pts=FALSE, fsize=0.8)
rate.mat.er<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
print(rate.mat.er)
```

What does this matrix mean?

```{r, eval=FALSE}
pp.er<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(pp.er) #lowerAIC
```

What do these results mean?

```{r, eval=FALSE}
rate.mat.ard<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ARD")
print(rate.mat.ard)
```

And these?

```{r, eval=FALSE}
pp.ard<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard,node.states="marginal")
print(pp.ard)
```

which model is better? 

Now let's look at multiple traits.

This is a matrix with four states

```{r, eval=FALSE}
rate.mat.er.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rate.mat.er.4state) #confused on what this is .. fake data? 
```

Convert the two binary traits into a single four character state

```{r, eval=FALSE}
fourstate.trait<-rep(NA,Ntip(primates$tree))
for(i in sequence(Ntip(primates$tree))) {
	if(primates$trait[i,2]==0 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-0
	}	
	if(primates$trait[i,2]==0 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-1
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-2
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-3
	}	
}
fourstate.data<-data.frame(Genus_sp=primates$trait[,1], T1=fourstate.trait)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, model="ER", node.states="marginal"))
print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat=rate.mat.er.4state, node.states="marginal", model="ARD"))
rate.mat.ard.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")
print(rate.mat.ard.4state)
```

Now let's make the equivalent of a GTR matrix:

```{r, eval=FALSE}
rate.mat.gtr.4state<-rate.mat.ard.4state
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(1,4))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(2,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(3,8))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(4,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(5,7))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(6,7))
print(rate.mat.gtr.4state)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.gtr.4state, node.states="marginal", model="ARD")) #even lower AIC #rate matrix! 
```


Now make a model like Pagel 1994

```{r, eval=FALSE}
print(rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=2, nstates=2, model="ARD"))
rate.mat.pag94<-rate.par.drop(rate.mat.ard.4state, drop.par=c(3,5,8,10))
print(rate.mat.pag94)
mod1<-corDISC(primates$tree,primates$trait,ntraits=2,model="ARD",node.states="marginal",diagn=FALSE)
print(mod1)
```

Now that you have some introduction, there are two routes:

##Route 1

**1). Construct a model to test if state 1 can never be lost**
```{r, eval=FALSE}
?corDISC #exploring this model and its limitations/flexibility
mod1a<-corDISC(primates$tree,primates$trait,ntraits=2,rate.mat=rate.mat.pag94,model="ARD",node.states="marginal",diagn=FALSE)
print(mod1) #aic 102.34
mod1b<-corDISC(primates$tree,primates$trait,ntraits=2,rate.mat=rate.mat.pag94,model="ER",node.states="marginal",diagn=FALSE)
print(mod2) 
mod1c<-corDISC(primates$tree,primates$trait,ntraits=2,rate.mat=rate.mat.pag94,model="SYM",node.states="marginal",diagn=FALSE) 

#no real substantial difference when I change the models above. I think I'll stick with ARD model

#Brian, what would be the potential limitation of trying to use fitPagel for modeling binary correlated traits? I've attempted to set some models up and just explore myself.. As if I were to complete the assignment using this function... 
#fitPagel #test model fit, AIC, indep/depend, biological conclusions. Can 1 be lost?  
#fit.t1t2<-fitPagel(primates$tree,primates$trait$T1,primates$trait$T2,dep.var="xy") #correlated #ARD
#fit.t1<-fitPagel(primates$tree,primates$trait$T1,primates$trait$T2,dep.var="x") #independent #ARD
#fit.t2<-fitPagel(primates$tree,primates$trait$T1,primates$trait$T2,dep.var="y") #independent #ARD
#fit.er<-fitPagel(primates$tree,primates$trait$T1,primates$trait$T2,model="ER") #equal rates, x has two rates depending on the state of y (but equal in forward and backwards rate)
#not sure what I'm up to but still messing around with fitPagel here.
#fit.g<-fitPagel(primates$tree,primates$trait$T1,primates$trait$T2,method="fitDiscrete") #geiger
#fit.er<-fitPagel(primates$tree,primates$trait$T1,primates$trait$T2,model="ER") #equal rates, x has two rates depending on the state of y (but equal in forward and backwards rate)
```
**2). Experiment with the effects of frequencies at the root.**
```{r, eval=FALSE} 
#messing around with root node prob. 
root.p2a<-c(1,0,0,0) #00,01,10,11
root.p2b<-c(0,1,0,0) #00,01,10,11
root.p2c<-c(0,0,1,0) #00,01,10,11
root.p2d<-c(0,0,0,1) #00,01,10,11
mod2a<-corDISC(primates$tree,primates$trait,ntraits=2,rate.mat=rate.mat.pag94,node.states="marginal",root.p=root.p2a,model="ARD",diagn=FALSE) #lowest AIC with fixed prob of state 00 at ancestral root
mod2b<-corDISC(primates$tree,primates$trait,ntraits=2,rate.mat=rate.mat.pag94,node.states="marginal",root.p=root.p2b,model="ARD",diagn=FALSE)
mod2c<-corDISC(primates$tree,primates$trait,ntraits=2,rate.mat=rate.mat.pag94,node.states="marginal",root.p=root.p2c,model="ARD",diagn=FALSE)
mod2d<-corDISC(primates$tree,primates$trait,ntraits=2,rate.mat=rate.mat.pag94,node.states="marginal",root.p=root.p2d,model="ARD",diagn=FALSE)
```
**3). Create and use a model to see if transitions from 00 go to 11 only via 01.**
```{r, eval=FALSE} 
#i think here I want to create an independent model with asymetrical transition rates, marginal node states to infer ancestral states at internal nodes, now how do I make one model independent and the other not with corDISC? Use rayDISC (Independence) and corDISC (dependence)
mod3a<-corDISC(primates$tree,primates$trait,ntraits=2,rate.mat=rate.mat.pag94,model="ARD",node.states="marginal",diagn=FALSE) #same as mod1a
#primT1<-c(primates$trait$Genus_sp,primates$trait$T1) attempting to isolate T1 and T2
#primT2<-c(primates$trait$Genus_sp,primates$trait$T2) to put in mod 3b and 3b1 to then sum the log lik of the independent models. 
mod3b<-rayDISC(primates$tree,primates$trait,model="ARD",node.states="marginal") #still running into errors when using primt1 or primt2. rayDISC only working with primates$trait
#attempting to then test model fit to see if a dependent model vs an independent model would be best. #loglikelihood test between summed independent log lik and the dependent log lik 
loglik3a<-mod3a$loglik #null
loglik3b<-mod3b$loglik #alt
ratio<-(loglik3a/loglik3b)
testa<-0.5*ratio
#testb<-2[(loglik3b)-(loglik3a)]
#reject null: but really close to 1... traits are equally as likely to be indep as correlated.

```
##Route 2

Maddison and FitzJohn (2015) pretty convincingly show (to me) that Pagel (1994) is just not a good method. Ok. So work on a fix. They point to [Read and Nee (1995)](http://dx.doi.org/10.1006/jtbi.1995.0047) as a low power but possible solution. Look at their appendix, especially, and write an implementation.
