
You can do this on your own data, or on included data here.


##Continuous data

```{r, eval=TRUE}
#install.packages("ape")
library(ape)
library(geiger)
tree.primates <- read.tree(text="((((Homo:0.21,Pongo:0.21):0.28,Macaca:0.49):0.13,Ateles:0.62):0.38,Galago:1.00);") #using examples from ape ?pic
X <- c(4.09434, 3.61092, 2.37024, 2.02815, -1.46968)
Y <- c(4.74493, 3.33220, 3.36730, 2.89037, 2.30259)
names(X) <- names(Y) <- c("Homo", "Pongo", "Macaca", "Ateles", "Galago")
pic.X <- pic(X, tree.primates)
pic.Y <- pic(Y, tree.primates)
```

Now, positivitize the contrasts and do a regression through the origin.

##Discrete data

```{r, eval=TRUE}
require("corHMM")
data(primates)
ls()
print(primates)
require(phytools)
```


Just to make this a better dataset for our exercise, I'm going to change some of the states (I want to have all four trait combinations present).
For actual analyses, of course, DO NOT INVENT YOUR DATA.

First, a review of discrete state models:

```{r, eval=TRUE}
primates$trait[which(grepl("Hylobates",primates$trait[,1])),2]<-1

trait1<-primates$trait[,2]
names(trait1)<-primates$trait[,1]
plotSimmap(make.simmap(primates$tree, trait1), pts=FALSE, fsize=0.8)
rate.mat.er<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
print(rate.mat.er)
```

What does this matrix mean?

```{r, eval=TRUE}
pp.er<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(pp.er)
```

What do these results mean?

```{r, eval=TRUE}
rate.mat.ard<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ARD")
print(rate.mat.ard)
```

And these?

```{r, eval=TRUE}
pp.ard<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard,node.states="marginal")
print(pp.ard)
```

which model is better?

one-parameter equal rates model (ER), a symmetric model (SYM) in which forwards and reverse transitions between states are constrained to be equal, and an all rates different matrix (ARD) where all possible transitions between states receive distinct parameters
lower AIC is better
er or ard
er is better because of the AIC score
simpler model given lower amount of data makes the equal parameters (ER) better for the given model.
so, pp.er is better


Now let's look at multiple traits.

This is a matrix with four states

```{r, eval=TRUE}
rate.mat.er.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rate.mat.er.4state)
```

Convert the two binary traits into a single four character state

```{r, eval=TRUE}
fourstate.trait<-rep(NA,Ntip(primates$tree))
for(i in sequence(Ntip(primates$tree))) {
	if(primates$trait[i,2]==0 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-0
	}	
	if(primates$trait[i,2]==0 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-1
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-2
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-3
	}	
}
fourstate.data<-data.frame(Genus_sp=primates$trait[,1], T1=fourstate.trait)

equal.rates.result <- rayDISC(primates$tree, fourstate.data, ntraits=1, model="ER", node.states="marginal")
print(rayDISC(primates$tree, fourstate.data, ntraits=1, node.states="marginal", model="ARD"))
print(rayDISC(primates$tree, fourstate.data, ntraits=1, node.states="marginal", rate.mat=rate.mat.er.4state, model="ARD"))
print(rate.mat.er.4state)
```

Now let's make the equivalent of a GTR matrix:

```{r, eval=TRUE}
rate.mat.gtr.4state<-rate.mat.er.4state
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(1,4))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(2,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(3,8))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(4,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(5,7))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(6,7))
print(rate.mat.gtr.4state)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.gtr.4state, node.states="marginal", model="ARD"))
```


Now make a model like Pagel 1994

```{r, eval=TRUE}
print(rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=2, nstates=2, model="ARD"))
rate.mat.pag94<-rate.par.drop(rate.mat.er.4state, drop.par=c(3,5,8,10))
print(rate.mat.pag94)
```

Now that you have some introduction, there are two routes:

##Route 1

**Construct a model to test if state 1 can never be lost**

**Experiment with the effects of frequencies at the root.**

**Create and use a model to see if transitions from 00 go to 11 only via 01.**

If state 1 can ever be lost
```{r}
rate.one.state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
rate.one.state[2,1]<-0
row.names(rate.one.state)<-c("0","1")
colnames(rate.one.state)<-c("0","1")
print(rate.one.state)
```

Effects of Frequency at the root
```{r}
root46<- corHMM(primates$tree, primates$trait, rate.cat=1, root.p = c(0.4,0.6))
print(root46)
root55<- corHMM(primates$tree, primates$trait, rate.cat=1, root.p=c(.5,.5))
print(root55)
```

With root46 there was a larger rate difference between the transitions from 0 to 1 than 1 to 0. With root55, the rates were fairly similar between transitions.

Create and use a model to see if transitions from 00 go to 11 only via 01
```{r}
rate.01<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
row.names(rate.01)<-c("00","01","10","11")
colnames(rate.01)<-c("00","01","10","11")
rate.01[2,1]<-1
rate.01[3,1]<-1
rate.01[4,1]<-0
rate.01[1,2]<-1
rate.01[3,2]<-0
rate.01[4,2]<-1
rate.01[1,3]<-0
rate.01[2,3]<-0
rate.01[4,3]<-1
rate.01[1,4]<-0
rate.01[2,4]<-1
rate.01[3,4]<-0
print(rate.01)
```



##Route 2

Maddison and FitzJohn (2015) pretty convincingly show (to me) that Pagel (1994) is just not a good method. Ok. So work on a fix. They point to [Read and Nee (1995)](http://dx.doi.org/10.1006/jtbi.1995.0047) as a low power but possible solution. Look at their appendix, especially, and write an implementation.